---
title: "Input Formats and Software Importers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input Formats and Software Importers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Supported software outputs

| Software | Version tested | Auto-detected | Import function |
|----------|---------------|---------------|-----------------|
| DIA-NN | ≥ 1.8 | ✓ | `import_diann()` |
| Spectronaut | ≥ 15 | ✓ | `import_spectronaut()` |
| FragPipe | ≥ 19 | ✓ | `import_fragpipe()` |
| MaxQuant | ≥ 2.0 | ✓ | `import_maxquant()` |
| Proteome Discoverer | ≥ 2.5 | ✓ | `import_pd()` |
| Skyline | any | ✓ | `import_skyline()` |
| Generic TSV/CSV | — | — | `import_generic()` |

## Auto-detection

```{r detect, eval=FALSE}
library(proteoforge)
sw <- detect_software("report.tsv")  # returns "diann", "spectronaut", etc.
```

`detect_software()` inspects filename patterns and column headers to identify
the originating software.

## DIA-NN

ProteoForge supports three DIA-NN output types:

```{r diann, eval=FALSE}
# 1. Protein-group matrix (pg_matrix.tsv)
pfd <- import_diann("pg_matrix.tsv", "samples.tsv")

# 2. Gene-level matrix (genes_matrix.tsv)
pfd <- import_diann("genes_matrix.tsv", "samples.tsv")

# 3. Full precursor report (report.tsv) — aggregates to protein level
pfd <- import_diann("report.tsv", "samples.tsv",
                     config = list(input = list(
                       diann_q_value = 0.01,
                       diann_min_peptides = 1L
                     )))
```

## Spectronaut

```{r spectronaut, eval=FALSE}
pfd <- import_spectronaut("spectronaut_export.tsv", "samples.tsv",
                           config = list(input = list(
                             spectronaut_q_cutoff = 0.01
                           )))
```

The importer uses `PG.Quantity` by default, falling back to `EG.TotalQuantity`.

## FragPipe

```{r fragpipe, eval=FALSE}
pfd <- import_fragpipe("combined_protein.tsv", "samples.tsv")
```

MaxLFQ columns are detected automatically (`MaxLFQ Intensity` suffix).
Proteins with `Probability < 0.9` or `< 1` unique peptide are filtered.

## MaxQuant

```{r maxquant, eval=FALSE}
# Protein groups
pfd <- import_maxquant("proteinGroups.txt", "samples.tsv")

# Phospho (STY) sites — expands multiplicity to separate rows
pfd <- import_maxquant("Phospho (STY)Sites.txt", "samples.tsv",
                        config = list(input = list(analysis_level = "phosphosite")))
```

## Proteome Discoverer

```{r pd, eval=FALSE}
pfd <- import_pd("proteins.txt", "samples.tsv",
                  config = list(input = list(
                    pd_sample_name_fn = function(x) sub("^Abundance: F\\d+: Sample, ", "", x)
                  )))
```

## Skyline

```{r skyline, eval=FALSE}
pfd <- import_skyline("skyline_report.csv", "samples.tsv")
```

Column auto-detection handles Protein Name, Replicate Name, and Total Area.

## Generic TSV/CSV

```{r generic, eval=FALSE}
pfd <- import_generic(
  quant_file     = "my_data.tsv",
  metadata_file  = "samples.tsv",
  feature_id_col = "ProteinID",
  config         = list(input = list(
    analysis_level = "protein",
    quant_type     = "LFQ"
  ))
)
```

Non-sample columns are inferred automatically from known patterns
(`ProteinID`, `GeneSymbol`, `Description`, etc.).

## Sample metadata format

The metadata file must be a TSV with at minimum:

```
sample_id   condition   batch
Sample_1    Control     B1
Sample_2    Control     B2
Sample_3    Treatment   B1
Sample_4    Treatment   B2
```

- `sample_id`: must match the column names in the quantification file
- `condition`: group labels for contrasts
- `batch` (optional): used for batch correction and design matrix

Use `validate_sample_metadata()` to check before importing.
