---
title: "Phosphoproteomics Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phosphoproteomics Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                       fig.width = 7, fig.height = 5)
```

## Overview

ProteoForge provides a dedicated phosphoproteomics pipeline that:

1. Filters to Class-1 localised sites
2. Standardises site identifiers
3. Corrects for parent-protein abundance changes
4. Runs differential abundance with limma-DEqMS
5. Infers kinase activity (KSEA, KEA3, decoupleR)
6. Runs phosphomotif enrichment (rmotifx)
7. Annotates sites with PhosphoSitePlus and OmniPath

## Loading phosphosite data

```{r import, eval=FALSE}
library(proteoforge)

quant_file <- system.file("extdata", "phospho_lfq_small",
                           "phospho_sites.tsv", package = "proteoforge")
meta_file  <- system.file("extdata", "phospho_lfq_small",
                           "sample_metadata.tsv", package = "proteoforge")

pfd_phospho <- import_generic(
  quant_file,
  meta_file,
  config = list(input = list(analysis_level = "phosphosite",
                              quant_type     = "LFQ")),
  feature_id_col = "SiteID"
)
```

## Step 1 — Localisation filter

```{r loc_filter, eval=FALSE}
# Retain Class-1 sites (loc_prob ≥ 0.75)
pfd_phospho <- filter_phospho_sites(pfd_phospho, loc_cutoff = 0.75)

# Standardise IDs to <UniProt>_<Residue><Position>
pfd_phospho <- standardise_site_ids(pfd_phospho,
                                     protein_col  = "ProteinID",
                                     residue_col  = "Residue",
                                     position_col = "Position")
```

## Step 2 — Preprocessing

```{r preprocess, eval=FALSE}
pfd_phospho <- filter_features(pfd_phospho)
pfd_phospho <- log2_transform(pfd_phospho)
pfd_phospho <- normalise_matrix(pfd_phospho, method = "median")
pfd_phospho <- impute_missing(pfd_phospho, method = "minprob")
```

For phosphosite data, MinProb imputation is preferred because most missing
phosphosites are MNAR (below-detection regulatory events).

## Step 3 — Parent-protein correction (optional)

Corrects phosphosite intensities for changes in total protein abundance,
isolating regulatory phosphorylation changes from stoichiometric ones.

```{r parent_correct, eval=FALSE}
# Load matched proteome
pfd_protein <- import_generic(
  system.file("extdata", "generic_lfq_small", "protein_intensities.tsv",
              package = "proteoforge"),
  meta_file
)
pfd_protein <- log2_transform(normalise_matrix(filter_features(pfd_protein)))

# Correct
correction_result <- parent_protein_correction(pfd_phospho, pfd_protein)
pfd_phospho_corrected <- correction_result$corrected_pfd
head(correction_result$correction_stats)
```

The regression model is: `log2(site) ~ log2(parent) + condition [+ batch]`

## Step 4 — Differential abundance

```{r de, eval=FALSE}
config  <- list(stats = list(contrasts = "EGF-Ctrl", fdr = 0.05, log2fc_cutoff = 0.5))
design  <- build_design_matrix(pfd_phospho_corrected, config)
cm      <- build_contrast_matrix(design, config)
ph_de   <- run_limma_deqms(pfd_phospho_corrected, design, cm, config)
```

## Step 5 — Kinase activity inference

Three methods are run and a consensus is applied (≥2 methods must agree at
adj.p < 0.05):

```{r kinase, eval=FALSE}
ka <- infer_kinase_activity(
  ph_de[["EGF-Ctrl"]],
  methods       = c("ksea", "decoupler"),
  consensus_min = 2L
)

# Per-method results
ka$ksea
ka$decoupler

# Consensus kinases
ka$consensus
```

KEA3 uses the public REST API (`https://maayanlab.cloud/kea3/`); an internet
connection is required. If unavailable, it is skipped with a warning.

## Step 6 — Phosphomotif enrichment

```{r motif, eval=FALSE}
motifs <- run_phospho_motif(
  ph_de[["EGF-Ctrl"]],
  window          = 7L,
  min_occurrences = 5L,
  p_cutoff        = 1e-6
)
head(motifs$up_motifs)
head(motifs$down_motifs)
```

## Step 7 — Site annotation

```{r annotate, eval=FALSE}
# OmniPath (free, no registration)
annotated <- annotate_phosphosites(
  ph_de[["EGF-Ctrl"]],
  use_omnipath = TRUE
)

# Add PhosphoSitePlus (requires user-downloaded files)
annotated <- annotate_phosphosites(
  ph_de[["EGF-Ctrl"]],
  psp_dir      = "~/psp_data",
  use_omnipath = TRUE
)
```

PSP data must be downloaded manually after registering at
`https://www.phosphosite.org/`.

## Running the complete pipeline automatically

```{r pipeline, eval=FALSE}
config <- yaml::read_yaml(
  system.file("config", "default_phospho.yml", package = "proteoforge")
)
config$input$quant_file    <- "phospho_sites.tsv"
config$input$metadata_file <- "samples.tsv"
run_pipeline(config)
```
