---
title: "ProteoForge Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProteoForge Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5
)
```

## Overview

ProteoForge is a complete R/Bioconductor proteomics analysis platform providing:

- Standardised importers for 6 major software outputs
- QC, normalisation, and imputation
- Limma-DEqMS differential abundance analysis
- Functional enrichment (ORA/GSEA) and STRING network analysis
- Phosphoproteomics with kinase activity inference
- Reproducible reporting (Excel, Word, PPTX, Quarto HTML)
- An interactive Shiny application

## Installation

```{r install, eval=FALSE}
# Bioconductor dependencies first
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("limma", "DEqMS", "QFeatures"))

# Install ProteoForge
devtools::install_github("proteoforge/proteoforge")
```

## Minimal end-to-end example

```{r example, eval=FALSE}
library(proteoforge)

# 1. Point to the bundled example dataset
quant_file <- system.file("extdata", "generic_lfq_small",
                           "protein_intensities.tsv", package = "proteoforge")
meta_file  <- system.file("extdata", "generic_lfq_small",
                           "sample_metadata.tsv",   package = "proteoforge")

# 2. Import
pfd <- import_generic(quant_file, meta_file,
                       config = list(input = list(analysis_level = "protein",
                                                   quant_type      = "LFQ")))

# 3. QC
qc <- compute_qc_metrics(pfd)
print(qc$outlier_samples)

# 4. Preprocess
pfd <- filter_features(pfd)
pfd <- log2_transform(pfd)
pfd <- normalise_matrix(pfd, method = "median")
pfd <- impute_missing(pfd, method = "mixed")

# 5. Differential abundance
config <- list(
  stats = list(
    contrasts     = "Treatment-Control",
    fdr           = 0.05,
    log2fc_cutoff = 1.0
  )
)
design   <- build_design_matrix(pfd, config)
contrasts <- build_contrast_matrix(design, config)
results  <- run_limma_deqms(pfd, design, contrasts, config)

# 6. Explore results
head(results[["Treatment-Control"]][significant == TRUE])

# 7. Volcano plot
plot_volcano(results[["Treatment-Control"]], contrast = "Treatment-Control")
```

## Interactive analysis

```{r shiny, eval=FALSE}
launch_app()
```

## Targets-based reproducible pipeline

```{r targets, eval=FALSE}
# Set config path and run
Sys.setenv(PF_CONFIG = "config.yml")
targets::tar_make()
```

## Getting help

- `?run_pipeline` — full pipeline documentation
- `?import_diann`, `?import_spectronaut` — software-specific importers
- `browseVignettes("proteoforge")` — all vignettes
