---
title: "Reproducibility and Deployment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducibility and Deployment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Reproducibility tools

ProteoForge provides four layers of reproducibility:

### 1. Parameter hashing

Every run computes a SHA-256 hash of the full configuration:

```{r hash, eval=FALSE}
library(proteoforge)
config <- yaml::read_yaml("config.yml")
hash   <- hash_params(config)
cat("Config hash:", hash, "\n")
# Writes hash to results/*/config_hash.txt
write_param_hash(config, results_dir = "results/run_001")
```

### 2. Package locking with renv

```{r renv, eval=FALSE}
renv::snapshot()    # lock current package versions
renv::restore()     # restore locked versions on a new machine
```

The `renv.lock` file is committed to the repository, ensuring byte-for-byte
reproducible package environments across collaborators and CI/CD systems.

### 3. Pipeline caching with targets

```{r targets, eval=FALSE}
# Only rerun steps whose inputs changed
targets::tar_make()

# Inspect which targets are outdated
targets::tar_outdated()

# Visualise the dependency graph
targets::tar_visnetwork()
```

The `_targets/` store caches each step's output. Unchanged intermediate
results (e.g., the import step) are never recomputed.

### 4. Session information

```{r session, eval=FALSE}
dirs <- make_output_dirs(config)
write_session_info(dirs$logs_dir)
```

Full `sessionInfo()` is written to `results/*/logs/session_info.txt` at the
end of every run.

## CLI flags

```bash
# Validate config and inputs, then exit
Rscript inst/scripts/proteoforge_cli.R \
  --config config.yml --validate-only

# Dry run: print planned steps without executing
Rscript inst/scripts/proteoforge_cli.R \
  --config config.yml --dry-run

# Resume (re-use cached targets from previous run)
Rscript inst/scripts/proteoforge_cli.R \
  --config config.yml --resume

# Regenerate reports from existing results
Rscript inst/scripts/proteoforge_cli.R \
  --config config.yml --report-only

# Run with 8 parallel workers
Rscript inst/scripts/proteoforge_cli.R \
  --config config.yml --cores 8
```

## Docker

### Analysis image

```bash
# Build
docker build -f docker/Dockerfile -t proteoforge:latest .

# Run pipeline
docker run --rm \
  -v $(pwd)/data:/data \
  -v $(pwd)/results:/results \
  proteoforge:latest \
  Rscript /app/inst/scripts/proteoforge_cli.R \
    --config /data/config.yml \
    --output-dir /results
```

### Shiny server image

```bash
docker build -f docker/Dockerfile.shiny -t proteoforge-shiny:latest .
docker run -d -p 3838:3838 proteoforge-shiny:latest
```

### Docker Compose (analysis + Shiny together)

```bash
docker-compose up
# Analysis UI: http://localhost:3838
# Results mounted to ./results/
```

## GitHub Actions CI

The repository includes four workflows:

| Workflow | Trigger | Purpose |
|----------|---------|---------|
| `R-CMD-check.yml` | push, PR | Full R CMD check on Ubuntu/macOS/Windows |
| `test-pipeline.yml` | push, PR | End-to-end pipeline with synthetic datasets |
| `lint.yml` | push, PR | lintr + styler style checks |
| `docker-build.yml` | push to main | Build and push Docker images to GHCR |

## Sharing analyses

The AI summary JSON (`ai_summary.json`) provides a machine-readable synopsis
of every analysis, enabling downstream LLM-assisted interpretation without
exposing raw data.

```{r ai, eval=FALSE}
# Read the summary
summary <- jsonlite::fromJSON("results/run_001/ai_summary.json")
cat("Significant proteins:", summary$contrast_summaries[[1]]$n_significant, "\n")
cat("Top up:", paste(summary$contrast_summaries[[1]]$top_up_proteins, collapse = ", "), "\n")
```
