---
title: "Using the ProteoForge Shiny Application"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the ProteoForge Shiny Application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Launching the app

```{r launch, eval=FALSE}
library(proteoforge)
launch_app()

# With parallel workers
launch_app(workers = 4L)

# For server deployment
launch_app(host = "0.0.0.0", port = 3838, launch = FALSE)
```

Or from the command line:

```bash
Rscript -e "proteoforge::launch_app()"
```

## Application layout

The app is organized as a tab-based navbar with 14 panels:

### 1. Welcome
Overview of capabilities and quick-start instructions.

### 2. Upload
- Select your input software (or Auto-detect)
- Upload quantification file and sample metadata
- Click **Import Data** — runs asynchronously so the UI stays responsive
- Preview the imported matrix

### 3. Column Mapping
Verify or correct automatic column assignments for gene symbols, protein IDs,
phosphosite localisation probabilities, and sequence windows.

### 4. Experimental Design
- Edit sample group assignments in the interactive table
- Define contrasts using the drop-down numerator/denominator builder
- Click **Add Contrast** to add multiple pairwise comparisons

### 5. QC
- Click **Compute QC Metrics** to generate all five QC plots:
  ID counts, intensity distributions, CV distribution, PCA, sample correlation
- Outlier samples are highlighted with a warning badge

### 6. Preprocessing
- Set filter thresholds (min valid fraction, min replicates)
- Choose normalisation method
- Choose imputation method
- Click **Run Preprocessing** — before/after boxplots update automatically

### 7. Statistics
- Set FDR and log₂FC thresholds
- Toggle DEqMS moderation
- Click **Run Statistics** — summary table with per-contrast counts
- Select a contrast to view the interactive volcano plot
- Download results as TSV

### 8. Heatmaps
Choose number of features, sorting criterion, contrast, and clustering options
to generate a publication-quality Z-scored heatmap.

### 9. Venn / UpSet
Select contrasts and direction (all/up/down) to generate Venn diagrams
(≤4 sets) or UpSet plots (≥5 sets), plus an intersection table.

### 10. Enrichment
Run ORA or GSEA on any contrast across GO/KEGG/Reactome databases. Results
are displayed as dot plots (ORA) and ridge plots (GSEA) with interactive tables.

### 11. Network
Run STRING network analysis on significant proteins from any contrast.
Export as PNG or GraphML for Cytoscape.

### 12. Phospho
Full phosphoproteomics sub-pipeline with localisation filtering, kinase
activity inference, and OmniPath/PSP annotation — available when a
phosphosite dataset is loaded.

### 13. Report
Generate all output formats:
- Excel workbook (all results + QC + enrichment)
- Word document with methods template
- PowerPoint slides with figures
- AI summary JSON
- Rendered Quarto HTML report

### 14. Logs
Live run log, warning list, and session information.

## Docker deployment

```bash
docker run -p 3838:3838 proteoforge/shiny:latest
```

See the [Docker guide](reproducibility.html) for full deployment instructions.
