---
title: "Statistical Methods in ProteoForge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Methods in ProteoForge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                       fig.width = 7, fig.height = 4.5)
```

## Preprocessing pipeline

### Filtering

`filter_features()` applies six filters in order:

1. Remove contaminants and decoy entries (MaxQuant `+` flag; generic patterns)
2. Require ≥ `min_valid_fraction` valid values in at least one condition
3. Require ≥ `min_replicates` valid values per condition
4. Remove all-NA rows
5. Apply minimum intensity threshold
6. Remove features with zero variance

### Normalisation

```{r norm, eval=FALSE}
library(proteoforge)
pfd <- normalise_matrix(pfd, method = "median")   # default
pfd <- normalise_matrix(pfd, method = "quantile")
pfd <- normalise_matrix(pfd, method = "vsn")
pfd <- normalise_matrix(pfd, method = "cyclic_loess")
```

All methods operate on log2-transformed intensities.

### Missingness classification

ProteoForge classifies missing values as **MNAR** (missing not at random) or
**MCAR** (missing completely at random) using a condition-level threshold:

- A site/protein is **MNAR** if ≥ 60% of samples in a condition are missing
  → implies the signal is genuinely below detection
- Otherwise **MCAR**

### Mixed imputation

```{r impute, eval=FALSE}
pfd <- impute_missing(pfd, method = "mixed")
# MNAR entries: MinProb (random draws from left tail of observed distribution)
# MCAR entries: kNN (k=5 nearest neighbours)
```

Other single-method options: `"minprob"`, `"mindet"`, `"knn"`, `"qrilc"`, `"rf"`.

## Differential abundance: limma-DEqMS

### Design and contrast matrices

```{r design, eval=FALSE}
config <- list(stats = list(contrasts = c("Treatment-Control", "DrugA-DrugB")))
design    <- build_design_matrix(pfd, config)
contrasts <- build_contrast_matrix(design, config)
```

`build_design_matrix()` creates a `~0 + condition` model; if a `batch` column
is present it is included as `~0 + condition + batch`.

### Fitting

```{r fit, eval=FALSE}
results <- run_limma_deqms(pfd, design, contrasts, config)
```

Steps:

1. `limma::lmFit()` — linear model per feature
2. `limma::contrasts.fit()` — extract specified contrasts
3. `limma::eBayes(trend=TRUE, robust=TRUE)` — empirical Bayes moderation
4. `DEqMS::spectraCounteBayes()` — peptide-count-based variance moderation
5. BH correction of p-values per contrast
6. Significance calls: `adj.P.Val < FDR & |log2FC| > threshold`

DEqMS requires peptide count information stored in `feature_metadata`. When
unavailable, the pipeline falls back to standard eBayes with a warning.

### Results structure

Each element of the returned list is a `data.table` with columns:

| Column | Description |
|--------|-------------|
| `feature_id` | Feature identifier |
| `GeneSymbol` | Gene name (if available) |
| `log2FC` | log₂ fold-change |
| `t` | moderated t-statistic |
| `P.Value` | raw p-value |
| `adj.P.Val` | BH-adjusted p-value |
| `significant` | `TRUE` if both FDR and FC thresholds met |
| `direction` | `"up"`, `"down"`, or `"ns"` |
| `pct_imputed` | fraction of values imputed per feature |

## Enrichment analysis

### ORA

```{r ora, eval=FALSE}
# Uses detected proteome as background universe (never whole genome)
ora_results <- run_ora(results[["Treatment-Control"]], pfd, config)
```

### GSEA

```{r gsea, eval=FALSE}
# Ranking: sign(log2FC) × -log10(adj.P.Val)
gsea_results <- run_gsea(results[["Treatment-Control"]], config)
```

Both use `clusterProfiler` internally with reproducible database versioning.

## Reproducibility

Every run records:

- SHA-256 hash of the full config: `write_param_hash()`
- Full session info: `write_session_info()`
- Timestamp-stamped output directories
- `renv.lock` for package version pinning
