---
title: "`r params$project_name`"
subtitle: "ProteoForge Quantitative Proteomics Report"
date: "`r params$analysis_date`"
author: "ProteoForge v`r params$proteoforge_version`"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-float: true
    code-fold: true
    self-contained: true
    fig-width: 8
    fig-height: 5
params:
  project_name: "ProteoForge Analysis"
  n_contrasts: 0
  contrast_names: []
  n_significant: 0
  n_proteins: 0
  software: "unknown"
  norm_method: "median"
  imputation_method: "mixed"
  fdr_threshold: 0.05
  fc_threshold: 1.0
  has_enrichment: false
  has_phospho: false
  analysis_date: ""
  proteoforge_version: "dev"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Executive Summary

This report summarises a label-free quantification (LFQ) proteomics analysis
performed with **ProteoForge** (v`r params$proteoforge_version`) on
`r params$n_proteins` quantified proteins across
`r params$n_contrasts` contrast(s).

| Parameter | Value |
|-----------|-------|
| Input software | `r params$software` |
| Normalisation | `r params$norm_method` |
| Imputation | `r params$imputation_method` |
| FDR threshold | `r params$fdr_threshold` |
| log₂FC threshold | `r params$fc_threshold` |
| Total significant | **`r params$n_significant`** |
| Analysis date | `r params$analysis_date` |

---

## Quality Control

```{r qc-plots, results='asis'}
qc_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "qc")
plots  <- list.files(qc_dir, pattern = "\\.png$", full.names = TRUE)
if (length(plots) > 0) {
  for (p in plots) {
    cat(sprintf('![QC plot](%s)\n\n', p))
  }
} else {
  cat("*QC plots not found in output directory.*\n")
}
```

---

## Differential Abundance Results

```{r contrast-tables, results='asis'}
de_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "de_results")
for (cname in params$contrast_names) {
  cat(sprintf("### %s\n\n", cname))
  tsv <- file.path(de_dir, sprintf("%s__significant.tsv", cname))
  if (file.exists(tsv)) {
    dt <- data.table::fread(tsv)
    print(knitr::kable(head(dt, 20), digits = 4,
                        caption = sprintf("Top significant hits: %s", cname)))
    cat("\n\n")
  }
  # Volcano plot
  vp <- list.files(file.path(dirname(knitr::current_input(dir = TRUE)), "plots"),
                   pattern = sprintf(".*volcano.*%s.*\\.png", gsub("[^A-Za-z0-9]", ".*", cname)),
                   full.names = TRUE)[1]
  if (!is.na(vp) && file.exists(vp)) {
    cat(sprintf('![Volcano: %s](%s)\n\n', cname, vp))
  }
}
```

---

```{r enrichment-section, results='asis', eval=params$has_enrichment}
cat("## Functional Enrichment\n\n")
enrich_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "enrichment")
ora_plots  <- list.files(enrich_dir, pattern = ".*dotplot.*\\.png$", full.names = TRUE)
for (p in ora_plots) cat(sprintf("![](%s)\n\n", p))
```

```{r phospho-section, results='asis', eval=params$has_phospho}
cat("## Phosphoproteomics\n\n")
phospho_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "phospho")
ph_plots <- list.files(phospho_dir, pattern = "\\.png$", full.names = TRUE)
for (p in ph_plots) cat(sprintf("![](%s)\n\n", p))
```

---

## Methods

```{r methods-ai-stub, results='asis'}
ai_json <- file.path(dirname(knitr::current_input(dir = TRUE)), "ai_summary.json")
if (file.exists(ai_json)) {
  summary <- jsonlite::fromJSON(ai_json)
  if (!is.null(summary$ai_insertion_points$methods_overview)) {
    cat(summary$ai_insertion_points$methods_overview, "\n\n")
  }
} else {
  cat(sprintf(
    "Label-free quantification data from **%s** were analysed using
ProteoForge v%s. Protein intensities were normalised by %s normalisation
and missing values were handled using a mixed MNAR/MCAR strategy (%s).
Differential abundance was assessed with limma-DEqMS (FDR < %s,
|log₂FC| > %s).\n\n",
    params$software, params$proteoforge_version,
    params$norm_method, params$imputation_method,
    params$fdr_threshold, params$fc_threshold
  ))
}
```

---

## Session Information

```{r session-info}
sessionInfo()
```

*Report generated by [ProteoForge](https://github.com/proteoforge/proteoforge).*
